### 2019-05-06 1-Preamble/4-Acknowledgement/acknowledgement.tex
So boring

I would like to acknowledge and thank the following people, and also acknowledge the various funding who supported our works during the last three years and made this thesis possible.

My first thoughts go to my two PhD advisors.
%
Firstly, thanks to \href{https://MoyChristophe.Wordpress.com/}{Christophe Moy} who welcomed me in the SCEE team, in IETR lab and in CentraleSupélec campus of Rennes, a very nice and warm environment for research and day-to-day life.
Christophe was very supportive on all aspects of the PhD work, and he was also helpful during difficult moments in the last three years. Many thanks for being so nice and dynamic\footnote{We had a nice first contact when we met in Spring 2015 in Mahindra Ecole Centrale, in Hyderabad, India, where I taught for one year, and I am honored to have pursued my PhD under his supervision and to have worked with him, and I am also happy to have lived great human moments along with Christophe since 2016.}, and always comprehensive.
Christophe was helpful for his expertise on wireless communications and engineering, and especially on cognitive radio.
Regardless of what we did together, whether it was debugging a difficult GNU Radio block \cite{Besson2018ICT,Besson2019WCNC}, writing a paper by night while Christophe was in a plane coming back from India \cite{MoyBesson2019}, chatting at a barbecue or playing ``palets'' together along the ocean in Vannes, I was lucky to collaborate with and be around Christophe.\\
%
\indent
Secondly, thanks to \href{http://chercheurs.lille.inria.fr/ekaufman/research.html}{Émilie Kaufmann}, who accepted to co-supervised my thesis, and did an amazing job in the last four years\footnote{Since our first discussions, in Autumn 2015 when she was giving practical sessions for the \href{http://researchers.lille.inria.fr/~lazaric/Webpage/MVA-RL_Course15.html}{Reinforcement Learning course} of \href{http://researchers.lille.inria.fr/~lazaric/}{Alessandro Lazaric} at MVA\footnote{\href{http://math.ens-paris-saclay.fr/version-francaise/formations/master-mva/}{Master Mathématiques Vision Apprentissage} at \href{https://www.ens-paris-saclay.fr/}{ENS de Paris-Saclay}, previously named \href{https://www.ens-cachan.fr}{ENS de Cachan}.}, to the last repetitions of my PhD presentation in October 2019.}.
Émilie was highly motivated by our collaboration, and she always pushed me further, to explore the mathematical aspects of my PhD work, to keep faith in the research world, and to keep the highest possible rigour in all parts of our exchanges and productions.
I was lucky to be able to visit the SequeL team at Inria Lille on regular occasions, and to work in close collaboration with Émilie during these visits\footnote{Merci à Romain et toi pour votre accueil chaleureux, et aussi merci à Florian, Nicolas et Édouard.}. Intensive white board sessions proved to be fruitful, giving the research work that I am the most proud of \cite{Besson2018ALT}, and rich research papers that join her rigorous expertise in statistical learning and mathematical analysis of bandit algorithms with my passion for numerical experiments, simulations and algorithm designs \cite{Besson2018ALT,Besson2018DoublingTricks,Besson2019GLRT}.

>>>>

I would like to acknowledge and thank the following people, and also acknowledge the various fundings who supported our works during the last three years and made this thesis possible.

My first thoughts go to my two PhD advisors.\\
%
\indent
Firstly, thanks to \href{https://MoyChristophe.Wordpress.com/}{Christophe Moy} who welcomed me in the SCEE team, in IETR lab and in CentraleSupélec campus of Rennes, a very nice and warm environment for research and day-to-day life.
Christophe was very supportive on all aspects of the PhD work, and he was also helpful during difficult moments in the last three years. Many thanks for being so nice and dynamic\footnote{We had a nice first contact when we met in Spring 2015 in Mahindra Ecole Centrale, in Hyderabad, India, where I taught for one year, and I am honored to have pursued my PhD under his supervision and to have worked with him, and I am also happy to have lived great human moments along with Christophe since 2016.}, and always comprehensive.
Christophe was helpful for his expertise on wireless communications and engineering, and especially on cognitive radio.
Regardless of what we did together, whether it was debugging a difficult GNU Radio block \cite{Besson2018ICT,Besson2019WCNC}, writing a paper by night while Christophe was in a plane coming back from India \cite{MoyBesson2019}, chatting at a barbecue or playing ``palets'' together along the ocean in Vannes, I was lucky to collaborate with and be around Christophe.\\
%
\indent
Secondly, thanks to \href{http://chercheurs.lille.inria.fr/ekaufman/research.html}{Émilie Kaufmann}, who accepted to co-supervised my thesis, and did an amazing job in the last four years\footnote{Since our first discussions, in Autumn 2015 when she was giving practical sessions for the \href{http://researchers.lille.inria.fr/~lazaric/Webpage/MVA-RL_Course15.html}{Reinforcement Learning course} of \href{http://researchers.lille.inria.fr/~lazaric/}{Alessandro Lazaric} at MVA\footnote{\href{http://math.ens-paris-saclay.fr/version-francaise/formations/master-mva/}{Master Mathématiques Vision Apprentissage} at \href{https://www.ens-paris-saclay.fr/}{ENS de Paris-Saclay}, previously named \href{https://www.ens-cachan.fr}{ENS de Cachan}.}, to the last repetitions of my PhD presentation in October 2019.}.
Émilie was highly motivated by our collaboration, and she always pushed me further, to explore the mathematical aspects of my PhD work, to keep faith in the research world, and to keep the highest possible rigour in all parts of our exchanges and productions.
I was lucky to be able to visit the SequeL team at Inria Lille on regular occasions, and to work in close collaboration with Émilie during these visits\footnote{Merci à Romain et toi pour votre accueil chaleureux, et aussi merci à Florian, Nicolas et Édouard.}. Intensive white board sessions proved to be fruitful, giving the research work that I am the most proud of \cite{Besson2018ALT}, and rich research papers that join her rigorous expertise in statistical learning and mathematical analysis of bandit algorithms with my passion for numerical experiments, simulations and algorithm designs \cite{Besson2018ALT,Besson2018DoublingTricks,Besson2019GLRT}.

### 2019-05-06 2-Chapters/1-Chapter/chapter1.tex
So boring

\paragraph{Tutorial}
    I gave a tutorial\footnote{Cf. \texttt{hal.archives-ouvertes.fr/cel-01830248}} on the Julia language, at IETR seminar in Vannes in June $2018$,
    with Pierre Haessig (\texttt{pierreh.eu}).

>>>>

\paragraph{Tutorial}
    I gave a tutorial\footnote{Cf. \href{HAL.Inria.fr/cel-01830248}{\texttt{HAL.Inria.fr/cel-01830248}}} on the Julia language, at IETR seminar in Vannes in June $2018$,
    with Pierre Haessig (\texttt{pierreh.eu}).

### 2019-05-06 2-Chapters/3-Chapter/chapter3.tex
So boring

\paragraph{List of algorithms}

@@

    \item Algorithm 9 is
    BESA (Best Empirical Sampled Average) from \cite{Baransi2014}, and it consists in trying to fix a simple observation.
    For instance in \UCB, if arm $1$ has $100$ samples and arm $2$ has $50$ samples, it should make more sense to only use $50$ samples of arm $1$ to compute any statistics for each arms (\eg, mean).
    The BESA algorithm for $K=2$ arms is then very simple: at each time step, if arm $1,2$ has $n_1,n_2$ samples and if $n_1>n_2$ (for instance), first sub-sample $n_2$ observations from $n_1$, then compute the mean of these observations, and play the arm maximizing this mean (now that there is the same number for both arms, intuitively it makes more sense to compare these quantities).
    It was analyzed partially for the two-armed case, and the authors illustrated that it can be very efficient empirically.
    Analyzing it in the generic case was found to be difficult, and is still an open problem.
    We implemented it using a binary tournament implement in a naive but efficient way (not using recursive functions) (\texttt{BESA} with \texttt{''non\_recursive''=True}).
    Due to its binary nature, the extension to more than $K=2$ arms is costly in terms of both time and memory, as illustrated below.
\end{itemize}

@@

% FIXME update if needed?
We consider increasing values for the horizon, and we focus on $7$ exponentially-spaced values: $T=1000$, $T=2500$, $T=5000$, $T=10000$, $T=20000$, $T=25000$ and $T=50000$.
We consider as well an increasing number of arms, and we focus again on $7$ values: $K=2$, $K=4$, $K=8$, $K=12$, $K=16$, $K=20$ and $K=32$.
For all experiments, we run $N=100$ independent simulations.

@@

            $R^3'_{T_1,K_1} = 0 \pm 0$
                $R^3'_{T_1,K_2} = 0 \pm 0$
                $R^3'_{T_1,K_3} = 0 \pm 0$
                $R^3'_{T_1,K_4} = 0 \pm 0$
                $R^3'_{T_1,K_5} = 0 \pm 0$ &
            $R^3'_{T_2,K_1} = 0 \pm 0$
                $R^3'_{T_2,K_2} = 0 \pm 0$
                $R^3'_{T_2,K_3} = 0 \pm 0$
                $R^3'_{T_2,K_4} = 0 \pm 0$
                $R^3'_{T_2,K_5} = 0 \pm 0$ &
            $R^3'_{T_3,K_1} = 0 \pm 0$
                $R^3'_{T_3,K_2} = 0 \pm 0$
                $R^3'_{T_3,K_3} = 0 \pm 0$
                $R^3'_{T_3,K_4} = 0 \pm 0$
                $R^3'_{T_3,K_5} = 0 \pm 0$ &
            $R^3'_{T_4,K_1} = 0 \pm 0$
                $R^3'_{T_4,K_2} = 0 \pm 0$
                $R^3'_{T_4,K_3} = 0 \pm 0$
                $R^3'_{T_4,K_4} = 0 \pm 0$
                $R^3'_{T_4,K_5} = 0 \pm 0$ \\

@@

\begin{figure}[h!]  % [htbp]
	% \centering
	\includegraphics[width=1.05\linewidth]{16_different_algorithms__lognormregret_vs_lognormtime__56pb__7KS_7Ts.pdf}
	\caption[Normalized mean regret vs normalized running time (in micro-seconds).]{
        Normalized mean regret vs normalized running time (in micro-seconds),
        aggregating the results from different values of $K$ and $T$, for $16$ algorithms.
        Both the $x$-axis and $y$-axis are in log-scale.
        \UCB{} and \klUCB{} are among the best algorithms, while AdBandits, Thompson sampling and Bayes-UCB slightly outperform them in terms of regret, and have similar running times.
	}
	\label{fig:3:16_different_algorithms__lognormregret_vs_lognormtime__56pb__7KS_7Ts}
\end{figure}

@@

\begin{figure}[h!]  % [htbp]
	% \centering
	\includegraphics[width=1.05\linewidth]{16_different_algorithms__lognormregret_vs_lognormmemory__56pb__7KS_7Ts.pdf}
	\caption[Normalized mean regret vs normalized memory costs (in bytes).]{
        Normalized mean regret vs normalized memory costs (in bytes),
        aggregating the results from different values of $K$ and $T$, for $16$ algorithms.
        Both the $x$-axis and $y$-axis are in log-scale.
        Thompson sampling appears as the best algorithm in this visualization, while \UCB{} has the advantage of being memory efficient, and \klUCB{} obtains similar performances.
	}
	\label{fig:3:16_different_algorithms__lognormregret_vs_lognormmemory__56pb__7KS_7Ts}
\end{figure}

@@

    \item Algorithm 13 is
    the anytime variant of KL-UCB-Switch from \cite{GarivierHadiji2018} (\texttt{klUCBswitchAnytime}).
    It is a very recent variant of the \klUCB{} algorithm \cite{KLUCBJournal},
    and it mimics the \klUCB{} algorithm but uses a much more complicated function to compute the upper confidence bounds, and uses more storage.
    It is anytime, and was proved to obtain good ``best of both world'' performances, meaning that it achieves an asymptotically optimal $\cO(\log(T))$ regret for problem-dependent bounds, and $\cO(\sqrt{KT})$ for problem-independent bounds.
    Empirically, it usually outperforms \klUCB{} and costs about the same memory, but it is more costly in terms of computation time.

@@

% https://tex.stackexchange.com/a/12430/
\begin{listing}[h!]
    \begin{minted}[linenos=true,numbersep=5pt,frame=lines,framesep=2mm]{bash}
$ for N in 4 100 1000;            do \
  for BAYES in False True;        do \
  for T in 1000 5000 10000 50000; do \
  for K in 2 5 10 20 50;          do \
    DEBUG=False NOPLOTS=False SAVEALL=True N_JOBS=-1 \
                         N=$N BAYES=$BAYES T=$T K=$K \
                    python3 main.py configuration.py \
    && cp logs/main_py3_log.txt \
          logs/main_py3_log__N${N}_BAYES${BAYES}_T{T}_K{K}.txt \
  done; done; done; done
    \end{minted}
    \caption{Snippet of Bash code to run the large experiments presented in Sections~\ref{sec:3:reviewSPAlgorithms} and \ref{sub:3:additionalExperiments}.}
    % \captionof{listing}{Small snippet of Bash code to run an experiment \label{lst:3:BashCodeToLaunchLargeExperiments}.}
    \label{lst:3:BashCodeToLaunchLargeExperiments}
\end{listing}

>>>>

% \paragraph{List of algorithms}

@@

    \item Algorithm 9 is
    BESA (Best Empirical Sampled Average) from \cite{Baransi2014}, and it consists in trying to fix a simple observation.
    For instance in \UCB, if arm $1$ has $100$ samples and arm $2$ has $50$ samples, it should make more sense to only use $50$ samples of arm $1$ to compute any statistics for each arms (\eg, mean).
    The BESA algorithm for $K=2$ arms is then very simple: at each time step, if arm $1,2$ has $n_1,n_2$ samples and if $n_1>n_2$ (for instance), first sub-sample $n_2$ observations from $n_1$, then compute the mean of these observations, and play the arm maximizing this mean (now that there is the same number of observations for both arms, intuitively it makes more sense to compare these quantities).
    It was analyzed partially for the two-armed case, and the authors illustrated that it can be very efficient empirically.
    Analyzing it in the generic case was found to be difficult, and is still an open problem.
    We implemented it using a binary tournament implement in a naive but efficient way (not using recursive functions) (\texttt{BESA} with \texttt{''non\_recursive''=True}).
    Due to its binary nature, the extension to more than $K=2$ arms is costly in terms of both time and memory, as illustrated below.
\end{itemize}

@@

% FIXME update if needed?
We consider increasing values for the horizon, and we focus on $7$ exponentially-spaced values: $T=1000$, $T=2500$, $T=5000$, $T=10000$, $T=20000$, $T=25000$ and $T=50000$.
We consider as well an increasing number of arms, and we focus again on $7$ values: $K=2$, $K=4$, $K=8$, $K=12$, $K=16$, $K=20$ and $K=32$.
For all experiments, we run $N=100$ independent simulations.
See the snippet of code~\ref{lst:3:BashCodeToLaunchLargeExperiments} in the Appendix below.

@@

            $R^{'3}_{T_1,K_1} = 0 \pm 0$
                $R^{'3}_{T_1,K_2} = 0 \pm 0$
                $R^{'3}_{T_1,K_3} = 0 \pm 0$
                $R^{'3}_{T_1,K_4} = 0 \pm 0$
                $R^{'3}_{T_1,K_5} = 0 \pm 0$ &
            $R^{'3}_{T_2,K_1} = 0 \pm 0$
                $R^{'3}_{T_2,K_2} = 0 \pm 0$
                $R^{'3}_{T_2,K_3} = 0 \pm 0$
                $R^{'3}_{T_2,K_4} = 0 \pm 0$
                $R^{'3}_{T_2,K_5} = 0 \pm 0$ &
            $R^{'3}_{T_3,K_1} = 0 \pm 0$
                $R^{'3}_{T_3,K_2} = 0 \pm 0$
                $R^{'3}_{T_3,K_3} = 0 \pm 0$
                $R^{'3}_{T_3,K_4} = 0 \pm 0$
                $R^{'3}_{T_3,K_5} = 0 \pm 0$ &
            $R^{'3}_{T_4,K_1} = 0 \pm 0$
                $R^{'3}_{T_4,K_2} = 0 \pm 0$
                $R^{'3}_{T_4,K_3} = 0 \pm 0$
                $R^{'3}_{T_4,K_4} = 0 \pm 0$
                $R^{'3}_{T_4,K_5} = 0 \pm 0$ \\

@@

\begin{figure}[h!]  % [htbp]
	% \centering
	\includegraphics[width=1.10\linewidth]{16_different_algorithms__lognormregret_vs_lognormtime__56pb__7KS_7Ts.pdf}
	\caption[Normalized mean regret vs normalized running time (in micro-seconds).]{
        Normalized mean regret vs normalized running time (in micro-seconds),
        aggregating the results from different values of $K$ and $T$, for $16$ algorithms.
        Both the $x$-axis and $y$-axis are in log-scale.
        \UCB{} and \klUCB{} are among the best algorithms, while AdBandits, Thompson sampling and Bayes-UCB slightly outperform them in terms of regret, and have similar running times.
	}
	\label{fig:3:16_different_algorithms__lognormregret_vs_lognormtime__56pb__7KS_7Ts}
\end{figure}

@@

\begin{figure}[h!]  % [htbp]
	% \centering
	\includegraphics[width=1.10\linewidth]{16_different_algorithms__lognormregret_vs_lognormmemory__56pb__7KS_7Ts.pdf}
	\caption[Normalized mean regret vs normalized memory costs (in bytes).]{
        Normalized mean regret vs normalized memory costs (in bytes),
        aggregating the results from different values of $K$ and $T$, for $16$ algorithms.
        Both the $x$-axis and $y$-axis are in log-scale.
        Thompson sampling appears as the best algorithm in this visualization, while \UCB{} has the advantage of being memory efficient, and \klUCB{} obtains similar performances.
	}
	\label{fig:3:16_different_algorithms__lognormregret_vs_lognormmemory__56pb__7KS_7Ts}
\end{figure}

@@

    \item Algorithm 13 is
    the anytime variant of KL-UCB-Switch from \cite{GarivierHadiji2018} (\texttt{klUCBswitchAnytime}).
    It is a very recent variant of the \klUCB{} algorithm \cite{KLUCBJournal},
    and it mimics the \klUCB{} algorithm but uses a much more complicated function to compute the upper confidence bounds, and uses more storage.
    It is anytime, and was proved to obtain good ``best of both world'' performances, meaning that it achieves an asymptotically optimal $\cO(\log(T))$ regret for problem-dependent bounds, and $\cO(\sqrt{KT})$ for problem-independent bounds.
    Empirically, it usually outperforms \klUCB{} in terms of regret, and costs about the same memory, but it is slower.

@@

% https://tex.stackexchange.com/a/12430/
\begin{listing}[h!]
    \begin{minted}[linenos=true,numbersep=5pt,frame=lines,framesep=2mm]{bash}
$ for N     in 4 100 1000;        do \
  for BAYES in False True;        do \
  for T in 1000 2500 5000 10000 20000 50000; do \
  for K     in 2 4 8 12 16 20 32; do \
    DEBUG=False NOPLOTS=False SAVEALL=True N_JOBS=-1 \
                         N=$N BAYES=$BAYES T=$T K=$K \
                    python3 main.py configuration.py \
    && cp logs/main_py3_log.txt \
          logs/main_py3_log__N${N}_BAYES${BAYES}_T{T}_K{K}.txt \
  done; done; done; done
    \end{minted}
    \caption{Snippet of Bash code to run the large experiments presented in Sections~\ref{sec:3:reviewSPAlgorithms} and \ref{sub:3:additionalExperiments}.}
    % \captionof{listing}{Small snippet of Bash code to run an experiment \label{lst:3:BashCodeToLaunchLargeExperiments}.}
    \label{lst:3:BashCodeToLaunchLargeExperiments}
\end{listing}

### 2019-05-06 3-Appendices/2-Appendix/appendix2.tex
So boring

This appendix quickly presents the contributions of another publication that was not presented in the main text of this thesis.
We studied doubling tricks and their possible use for multi-armed bandits, between fall 2017 and spring 2018, and we wrote an article \cite{Besson2018DoublingTricks} which was rejected at the COLT 2018 conference, and unfortunately we lacked the time to improve it and resubmit it to another conference.

@@

To conclude this work, we would need to complete the study of the doubling tricks, and instead of focussing on two families of doubling tricks (geometric and exponential), we need to completely characterize the doubling tricks that allow to preserve certain regret bound.
Such doubling is given either by the function mapping the current time $t$ to the current estimate of the horizon $T(t)$, or the sequence $(T_i)_{i\in\N^*}$ of successive estimates of the horizons.
We are interested in pursuing this work, in the hope of finding an intermediate sequence, growing faster than a geometric but slower than an exponential doubling sequence, that can preserve both worlds, problem-dependent bounds in $\cO(\log(T))$ and problem-independent bounds in $\cO(\sqrt{T})$.

>>>>

This appendix quickly presents the contributions of another publication that was not presented in the main text of this thesis.
We studied doubling tricks and their possible uses for multi-armed bandits, between fall 2017 and spring 2018, and we wrote an article \cite{Besson2018DoublingTricks} which was rejected at the COLT 2018 conference, and unfortunately we lacked the time to improve it and resubmit it to another conference.

@@

To conclude this work, we would need to complete the study of the doubling tricks, and instead of focussing on two families (geometric and exponential), we need to completely characterize the doubling tricks that allow to preserve certain regret bound.
Such doubling is given either by the function mapping the current time $t$ to the current estimate of the horizon $T(t)$, or the sequence $(T_i)_{i\in\N^*}$ of successive estimates of the horizons.
We are interested in pursuing this work, in the hope of finding an intermediate sequence, growing faster than a geometric but slower than an exponential doubling sequence, that can preserve both worlds, problem-dependent bounds in $\cO(\log(T))$ and problem-independent bounds in $\cO(\sqrt{T})$.
